---
output: github_document
---


```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "img/README-pointbin-",
  out.width = "100%"
)
```

# Geom `geom_pointbin`

The R package [ggpointdensity](https://github.com/variani/ggpointdensity) 
implements two methods for 2D density estimation for the `geom_pointdensity` geom, 
nearest neighbors (`method="default"`) and 
a kernel density approach `kde2d` (`method="kde2d"`) from the MASS package.
These two methods can be rather slow when datasets  are large (>100,000 observations)
and/or the disribution is not Gaussian.

A new geom `geom_pointbin` is based on the average shifted histogram 
implemented in the R package [ash](https://cran.r-project.org/web/packages/ash/)
and found to be both accurate and computationally efficient
[(Deng and Wickham, 2011)](https://vita.had.co.nz/papers/density-estimation.pdf).

```{r inc}
library(ggpointdensity)

library(ggplot2)
library(dplyr)
library(viridis)

library(cowplot)
theme_set(theme_cowplot(12))
```

## `geom_pointbin` accuracy

```{r sim_data}
N = 1e3
dat = bind_rows(
  tibble(x = rnorm(N, sd = 1), y = rnorm(N, sd = 10), group = "foo"),
  tibble(x = rnorm(N, mean = 1, sd = .5), y = rnorm(N, mean = 7, sd = 5), group = "bar"))
```

```{r plot_pointdensity_vs_pointbin}
p_base = ggplot(data = dat, mapping = aes(x = x, y = y))

p_nn = p_base + 
  geom_pointdensity(method = "default") +
  scale_color_viridis()

p_kde = p_base + 
  geom_pointdensity(method = "kde2d") +
  scale_color_viridis()

p_bin = p_base + 
  geom_pointbin() +
  scale_color_viridis()

p_bin2 = p_base + 
  geom_pointbin(nbin = 16) +
  scale_color_viridis()

plot_grid(
  p_nn + labs(subtitle = 'geom_pointdensity(method = "default")'),
  p_kde + labs(subtitle = 'geom_pointdensity(method = "kde2d")'),
  p_bin + labs(subtitle = 'geom_pointbin()'),
  p_bin2 + labs(subtitle = 'geom_pointbin(nbin = 16)'))
```

## Support of ggplot2 features

```{r facet}
p_bin + facet_wrap( ~ group)
```

```{r xlim}
p_bin + xlim(c(-3, 0))
```

## `geom_pointbin` computational efficiency

```{r data_diamonds}
data(diamonds, package = "ggplot2")
```

```{r plot_diamonds}
p_diamonds_base = ggplot(diamonds, aes(x = carat, y = price))
p_diamonds_bin = p_diamonds_base + 
  geom_pointbin() + scale_color_viridis()
ggplot2::benchplot(p_diamonds_bin)
```

```{r plot_diamonds_kde}
p_diamonds_kde = p_diamonds_base +
  geom_pointdensity(method = "kde2d") + scale_color_viridis()
ggplot2::benchplot(p_diamonds_kde)
```

```{r plot_diamonds_nn}
p_diamonds_nn = p_diamonds_base +
  geom_pointdensity(method = "default") + scale_color_viridis()
ggplot2::benchplot(p_diamonds_nn)
```

```{r plot_diamonds_bin2}
# default nbin = 64
p_diamonds_nbin64 = p_diamonds_base + 
  geom_pointbin() + scale_color_viridis()
p_diamonds_nbin128 = p_diamonds_base + 
  geom_pointbin(nbin = 128) + scale_color_viridis()

plot_grid(
  p_diamonds_nbin64, 
  p_diamonds_nbin128,
  ncol = 1)
```

## `geom_pointbin` sub-sampling within bins

```{r nsample_bin}
# 0 samples per bin
p_bin_samp0 = p_base + 
  geom_pointbin(nbin = 16, nsamples = 0) +
  scale_color_viridis()

# 100 samples per bin
p_bin_samp100 = p_base + 
  geom_pointbin(nbin = 16, nsamples = 100) +
  scale_color_viridis()

# 5 samples per bin
p_bin_samp5 = p_base + 
  geom_pointbin(nbin = 16, nsamples = 5) +
  scale_color_viridis()

# 1 sample per bin
p_bin_samp1 = p_base + 
  geom_pointbin(nbin = 16, nsamples = 1) +
  scale_color_viridis()

plot_grid(
  p_bin + labs(subtitle = "nbin = 16, nsamples = 0 (no sampling)"),
  p_bin_samp100 + labs(subtitle = "nbin = 16, nsamples = 100 (representative sampling)"),
  p_bin_samp5 + labs(subtitle = "nbin = 16, nsamples = 5 (very few samples per bin)"),
  p_bin_samp1 + labs(subtitle = "nbin = 16, nsamples = 1 (only one sample per bin)"),
  ncol = 2)
```

